\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{upgreek}
\usepackage{float}
\usepackage{graphicx}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}




\begin{document}

\title{Important information on the Physical Chemistry laboratory practice for pharmacy students}
%\author{Andr\'{a}s Kiss, assistant lecturer}
%Department of General and Physical Chemistry, Faculty of Sciences, University of P\'{e}cs, 7624 P\'{e}cs, Ifj\'{u}s\'{a}g \'{u}tja 6, Hungary
\date{2016.09.06}
\maketitle
%\address[akiss, gnagy]{J\'{a}nos Szent\'{a}gothai Research Centre, University of P\'{e}cs, 7624 P\'{e}cs, Ifj\'{u}s\'{a}g \'{u}tja 20, Hungary}
%\ead{akiss@gamma.ttk.pte.hu}

The semester consists of 12 laboratory practices, and two tests. For each practice, thorough preparation is recommended. Before each practice, you must pass a small entrance test, answering a few questions about the topic. If you cannot answer these questions sufficiently, you cannot start the practice (failed practice), and have to do a do-over on the 14th week. During the practice, you must record the measured data in your laboratory notebook with a non-erasable pen. Write only in your notebook, and not on separate pieces of paper. First and foremost, your notebook should be up-to-date and complete, and then pretty. If you make corrections, cross the incorrect entry with a single line, so as to be able to read it afterwards. At the end of the practice I will sign your data. Please don't leave without it, because you will not be able to prove you used your own data.

Additionally, after finishing the practice, you must evaluate the data, calculate the required results (at home), and submit it by an agreed deadline every week. You must use your own data, and no-one else's. Each practice's notebook will be graded on a 0 - 5 point scale. At the end of the semester, a total of 60 points can be gathered from notebooks. 

You will write the two tests: on the 7th, and on the 13th week. You can gather 12-12 points from these tests by answering the questions for all 6-6 practices, including the one you will do on the day of the test.

You need to reach at least 6 points (50\%) in both tests to finish the course. You also need to have at least 1 point for each practice to have it accepted. If you do the practice, but fail to evaluate it, you will get zero points for that practice.

You can miss one practice, but then you have to make up for it with a do-over on the 14th week. You can re-write the tests as well.

To summarize: 

\begin{enumerate}
\item 60 points from practices + notebooks (12 $\times$ 5 points). Need to have at least 1 point for each to pass.
\item 12 points from the first test. At least 50\% necessary to pass.
\item 12 points from the second test. At least 50\% necessary to pass. 
\end{enumerate}

If you meet all these requirements, you will get a final grade based on the sum of your points:

\begin{center}
\begin{tabular}{|r|l|}
\hline
points & grade \\
\hline
0 -- 41 & 1 (insufficient) \\
\hline
42 -- 53 & 2 (satisfactory) \\
\hline
54 -- 63 & 3 (average) \\
\hline
64 -- 73 & 4 (good) \\
\hline
74 -- 84 & 5 (excellent) \\
\hline
\end{tabular}
\end{center}

The tests will be written on the Friday of the 7th week (2016.10.21) and the 13th week (2016.12.02) 16:00. Do-over for one practice and the two tests: Friday the 14th week (2016.12.09) 16:00. 

Notebook submission: before the agreed deadline in the glassdoor cupboard in the corridor of the Department of General and Physical Chemistry, Ifjúság útja 6, ground floor. After correction and grading, you can pick up your notebook to prepare for the next practice.

Take the deadline seriously, as you will receive penalty for not submitting the notebook on time.

\vspace{10mm}
%\vfill

András Kiss

Assistant lecturer

Department of General and Physical Chemistry

Faculty of Sciences

University of Pécs

H-7624 Pecs, Ifjúság útja 6.

room: A/115 106

e-mail: akiss@gamma.ttk.pte.hu

university extension: +36 72 501 500 / 61021

(www.ttk.pte.hu)

\end{document}
